Building Suitable Datasets for Soft Computing and Machine Learning Techniques from Meteorological Data Integration: A case study for predicting significant wave height

### 2020-11-17 latex_manuscript/MDPI/2_energies_special_issue/template.tex
Letters for Speccial Issue Energies

%=================================================================
% Full title of the paper (Capitalized)
\Title{\textcolor{red}{Building Suitable Datasets for Soft Computing and Machine Learning Techniques from Meteorological Data Integration: A case  study for predicting significant wave height and flux of energy}}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Meteorological data are extensively used to perform environmental learning. \textcolor{red}{Soft Computing (SC) and} Machine Learning (ML) techniques, a valuable support in many research areas, require datasets containing information related to the topic under study, which are not always available in an appropriate format, and its preparation and pre-processing implies a lot of time and effort by the researchers. This paper presents a novel software tool with an user-friendly GUI to create datasets by means of management and data integration of meteorological observations from two well-known sources of information: the \textit{National Data Buoy Center} and the \textit{National Centers for Environmental Prediction and for Atmospheric Research Reanalysis Project}. Such datasets can be created using buoys and reanalysis data by customisable procedures, in terms of predictive and objective variables and temporal resolution. These datasets can be used by \textcolor{red}{SC and} ML methodologies for prediction tasks (classification or regression) that support improvement of sustainability energy production, design of production systems as WECs or environmental modelling, among others.}

% Keywords
\keyword{Environmental Prediction; \textcolor{red}{Renewable Energy Resource Evaluation}; Meteorological Data; Reanalysis Data; Marine Energy;   \textcolor{red}{Suitable Datasets for Soft Computing Models}}  % List three to ten pertinent keywords specific to the article, yet reasonably common within the subject discipline.

@@

		A better understanding of the environment is of vital importance for science, contributing not only to more efficient exploitation of natural resources but also to the development of new strategies aimed at its protection. In that sense, meteorological observations provide an essential and valuable source of information which is widely used by researchers to address environmental learning, comprehension, prediction and conservation in numerous oceanic and atmospheric studies of a wide variety of areas (e.g. energy, climate change, agriculture, etc.). Some specific examples of the diversity of fields in which meteorological data can be used in are, among others: estimation of global solar radiation based on sunshine duration \cite{SHAHRUKHANIS2019179}, directional analysis of sea storms \cite{LAFACE201545}, \textcolor{red}{estimation of hybrid energy systems taking into account economic and environmental objectives\cite{Kumar2020}}, wind power ramp events prediction \cite{DORADOMORENO2017428}, \textcolor{red}{sea surface temperature prediction \cite{He2020}}, study of the responses exhibited by plankton to fluid motions \cite{FUCHS2016109}, trends in solar radiation \cite{SILVA20101852} or simulation of extreme near shore sea conditions \cite{GOULDBY201415}. All these studies require a prior data collection and its adaptation to a specific format that allows the interpretation of them.

		\textcolor{red}{Once good quality and well-formed data are obtained, these can be used to extract information and building prediction models that explain the behavior of a certain problem. Although Soft Computing (SC) and Machine Learning (ML) techniques have the ability to handle uncertainty in data and are extensively used for modeling purposes, the real challenge in modelling studies is due to the inadequacy of data, since the adequacy of the models depend mainly on the the quality of the information used, so that if a researcher does not have quality data there will be no quality models.}

		\textcolor{red}{Continuing with this line}, special purpose software is usually developed to help researchers to advance in their studies related to energy and environmental modelling, becoming a great support for decision-making in the exploitation and protection of the environment. \textcolor{red}{In \cite{Neeraj2020} a software package in R called ``ForecastTB'' is developed to compare the accuracy of different forecasting methods as related to the characteristics of a time series dataset, presenting the software as a stepping stone in ML automation modeling.}
		%In \cite{CAMARGONOGUEIRA2016361}, a software tool for designing solar water heating systems is developed, which simulates different situations and finds the best technical and economical solution. 
		In \cite{LO2015293}, an integrated simulation tool for the optimum design of bifacial solar panel with reflectors is presented. This tool can also be used to analyse the performance of the solar cells. A framework for data integration of offshore wind farms is implemented in \cite{NGUYEN2013150} in order to facilitate data exchange and improve operation and maintenance practices. 
		\textcolor{red}{In \cite{Roberta2020} a new software named ``Storage LCA Tool`` is presented as tool for comparing PCM - phase change materials- storage systems  with traditional systems that do not entail energy storage, being beneficial in order to support decision-making on energy concepts for buildings.}
		%In \cite{SHEHADEH2020100644}, a load management application to maximize the used energy of thermal systems and reduce the costs and emissions is developed. 
		A risk assessment tool to improve safety standards and emergency management in onshore wind farms, is presented in \cite{ASTIASOGARCIA201648}. Raabe et al. \cite{RAABE2010213} developed two software tools, \textit{Model of Equilibrium of Bay Beaches} (MEPBAY) and \textit{Coastal Modelling System} (SMC), to support different operational levels of headland-bay beach in coastal engineering projects, and Motahhir et al. \cite{MOTAHHIR20199} developed an open hardware/software test bench for solar tracker.

		Marine energy prediction is currently a hot topic where meteorological data is used in. Marine Renewable Energy (MRE) is one of the most important renewable and sustainable energy sources available in our environment \textcolor{red}{\cite{en12050787}}, and it includes ocean thermal energy, marine tidal current energy and wave energy, among others. Its benefits and great potential \cite{ZEYRINGER20181281} make it one of the most relevant natural resources, playing a crucial role not only in the reduction of the emission of greenhouse gases but also in all other aspects involved in the difficult challenge of the transition to a low carbon footprint society \textcolor{red}{\cite{en12091657, BREDE201344}}. Wave energy exhibits a more stable power supply than wind energy and even solar energy. In recent years \textit{Wave Energy Converters} (WECs) \cite{FALCAO2010899} have been developed and widely installed to transform this wave energy into electricity, which can be injected into the electric network or supplied to existing offshore oil and gas platforms \cite{OLIVEIRAPINTO2019556} or seawater desalination plants \cite{FERNANDEZPRIETO2019546}, among others. WECs are mechanical devices that convert kinetic energy into electrical energy by means of either the vertical oscillation of waves or the linear motion of them. Nevertheless, waves are difficult to be characterised due to their stochastic nature, because of the influence of a large number of environmental factors that exert on them \cite{ochi1998}. As a consequence of this complexity, many aspects of WEC design, deployment and operation \cite{CROWLEY2018159, Abdelkhalik2016, 6898109} need a proper prediction of waves, in order to maximise the wave energy extraction \textcolor{red}{\cite{en80910370}}. For this purpose WECs use wave \textit{flux of energy} ($F_e$) which can be calculated from the two most important wave parameters in this regard: \textit{significant wave height} ($H_s$) and \textit{wave energy period} ($T_e$). Additionally, wave predictions \textcolor{red}{\cite{en11010011}} are also helpful for designing offshore structures \cite{CHATZIIOANNOU2017126}, operational works in the sea \cite{DALGIC2015211}, providing technical information to the coastal and coral reef planners in developing countries \cite{CALLAGHAN2018123}, etc.

		Currently, and as a support to traditional study procedures, \textcolor{red}{SC and ML techniques} \textcolor{red}{\cite{Sang-Yong2014,Bishop:2006:PRM:1162264}} are being widely used in numerous research fields related to classification, regression and optimisation tasks, obtaining significant improvements in the performance of the results \textcolor{red}{\cite{Fi-John2019,Mosavi2019,GUO201816}}. \textcolor{red}{SC and} ML methodologies can be used not only by experienced computer scientists but also by other researchers. For example, the well-known \textit{Waikato Environment for Knowledge Analysis} (WEKA) \cite{WEKA} software tool provides researchers with a wide collection of ML  algorithms. In this way, ML techniques have been already applied to tackle wave characterisation, accurately estimating $H_{s}$ and $T_{e}$ parameters \cite{DURANROSAL2017268, KUMAR2017605}, given that robustness of ML methods can tackle the previously explained difficulties in wave energy prediction. The problem is that, in order to apply these methods, it is essential to obtain datasets with relevant information about the issue under study, used to infer knowledge. Usually, these datasets are not publicly available in a friendly format, and their generation is the first step needed.

		\textcolor{red}{VOY POR AQUI!!!!!!}

		The information to create these datasets can be obtained from meteorological observations, but such information may be available in a inappropriate format and even contain missing values or measurements. Consequently, it is usually required to perform pre-processing tasks for improving the quality of the data, such as the replacement of missing values, outlier detection or data normalisation, among others. Furthermore, if more than one source of information is used to achieve a better characterisation of the problem under study \cite{JOHANSSON2015143, FERNANDEZ201544, Adams2010}, then a data integration process, denominated as the matching process in this document, has to be carried out by the researchers to manually create the datasets with the needed information. Moreover, depending on the subject and the ML technique to be applied, or even if the researcher considers other factors in order to improve the results or have more in-depth conclusions, the datasets would have to be updated afterwards. In summary, many important details and different intermediate steps have to be considered when creating suitable datasets for ML techniques, resulting in an extremely tedious task.

		The main purpose of this paper is to present a new open source tool for the creation of datasets integrated by meteorological variables from two sources of information. Given that the tool provides an user-friendly graphical interface, no knowledge in programming languages is needed, and it also prevents researchers from performing the mentioned tedious work and greatly simplify all the steps involved in it, avoiding possible errors in the intermediate steps. The meteorological data used by the tool come from two well-known sources of information: the \textit{National Oceanic and Atmospheric Administration} (NOAA) \textit{National Data Buoy Center} (NDBC) \cite{NOAA} and the \textit{National Centers for Environmental Prediction} (NCEP)/\textit{National Center for Atmospheric Research} (NCAR) \textit{Reanalysis Project} (NNRP or R1) \cite{Kalnay1996, Kistler2001}. The software tool presented in this work is named SPAMDA (Software for Pre-processing and Analysis of Meteorological DAta to build datasets). As SPAMDA performs all this data processing, it reduces the time involving these tasks and allows the researchers focus on the study of the meteorological aspects of the observations. The datasets obtained are ready to be used as input for ML techniques in prediction tasks (classification or regression), although the researchers can use them for other purposes. These datasets contain one or more meteorological variables as inputs and one variable as target (variable to be predicted). The format of the generated datasets will be \textit{Attribute-Relation File Format} (ARFF) \cite{WEKA_ARFF}, which is the one used by WEKA. Besides, the datasets can also be generated in \textit{Comma-Separated Values} (CSV) format, enabling the researchers to use others tools.

		Up to now, and to the best of our knowledge, this is the first software tool addressing the problem previously discussed, integrating meteorological data from NDBC and NNRP. The novelties and functionalities that SPAMDA offers to the researchers will be detailed in Section \ref{sec:SPAMDA}, although some of them are briefly summarised below \textcolor{red}{Esto hay que revisarlo}:
		\begin{itemize}
			\item The generation of datasets becomes a very easy and customisable task, by means of the selection of different input parameters, such as predictive and objective variables, classification and regression ML techniques, output discretisation or prediction horizon, among others.
			\item It makes the researcher focus on environmental modelling, without having to worry about mechanical tasks, avoiding laborious pre-processing procedures that imply a lot of time and effort. \textcolor{red}{Decir que permite un estudio previo o sondeo antes de hacer estudios más exhaustivos}
			\item It manages the extensive casuistry of data integration which can lead to incomplete datasets.
			\item It avoids possible researcher errors in the intermediate steps of the process, such as time and geographical coordinates conversion or missing values handling (dates or measurements not recorded), among others.
			\item It provides information about the quality and quantity of the data.
			\item It includes different pre-processing tasks, such as normalisation and missing data recovery.\textcolor{red}{Hay que quitarle importancia al tema del preprocesamiento, para que no nos comparen con eso.}
			\item It facilitates data management and well-organised storage of the datasets.
			\item The created datasets can be easily used by ML tools.
			\item Its modular design allows the implementation of new modules for managing meteorological data from others sources, benefiting future renewable energy and environmental research.
			\item It includes a user-friendly GUI, facilitating and greatly simplifying data management, and it is integrated with the Explorer environment of WEKA.
			\item It is multi-platform, and it can be used on any computer with Java regardless of the operating system.
		\end{itemize}

@@

				\textcolor{red}{Aquí en esta subsección hay que intentar darle poca importancia respecto a que el software no se centra en los filtros de preprocesamiento}

				Data pre-processing prepares the raw data (intermediate datasets) to be able to be treated correctly by ML algorithms. In this way, the quality of dat can be improved prior to the learning phase, by applying pre-processing tasks (filters). The result will be referenced as pre-processed datasets.

@@

@@

				In order to use the data provided by the two sources of information described in Section \ref{sec:DataSources}, it is necessary to carry out a data integration process, denominated as the matching process in this document, to merge and format such data. The matching procedure is performed using an intermediate or pre-processed dataset, which includes the measurements collected by a buoy from NDBC, and the needed reanalysis data files from NNRP. Note that SPAMDA is able to manage the NetCDF binary format for handling the information stored in the reanalysis files.

@@

							\item \textit{One (using weighted mean of the $N$ nearest reanalysis nodes)}: Only one final dataset will be created, which will contain the attributes (the selected one as output and the selected ones as inputs) of the intermediate or pre-processed dataset used, along with a weighted mean of each variable of the reanalysis data used (one per selected reanalysis data file). This weighted mean is obtained by SPAMDA and uses Eq. \ref{eq:Haversine} to obtain the distance from each reanalysis node to the location of the buoy. Once the distances have been calculated they are inverted and normalised as follows:							
								\begin{linenomath*}
									\begin{equation}
										w_i=\frac{\sum_{j=1}^{N} d(p_0,p_j)}{d(p_0,p_i)}, ~~i=1, \ldots, N.
										\label{eq:weightedMean}
									\end{equation}
								\end{linenomath*}

@@

							\item \textit{'N' (one per each reanalysis node)}: As many final datasets as the number of nearest $N$ reanalysis nodes configured by the researcher will be created. Therefore, each final dataset will contain the value of each reanalysis variable used of the nearest corresponding reanalysis node, along with the selected attributes of the intermediate or pre-processed dataset used.

@@

	%\section{Case study}\label{sec:CaseStudy}
	\section{\textcolor{red}{A case study applied to Gulf of Alaska}}\label{sec:CaseStudy}		
		This section will describe how the application works in a practical approach. To do so, an example showing how to create a fully processed dataset (final dataset), starting from the raw data, will be described. The objective of this final dataset is to be used with ML algorithms for environmental modelling, concretely, to classify waves in the Gulf of Alaska depending on their height. In this case, the problem will be tackled as a multi-class classification problem, given that any continuous variable can be discretised in different classes. Such waves modelling can be applied with different purposes, energy flux range forecasting, missing buoy data reconstruction, or extreme significant wave heights detection, among others.

@@

			After gathering the information described above\footnote{Further instructions for downloading this data can be found in the user manual of the application.}, the researcher can open SPAMDA. In Figure \ref{fig:main_view}, the main window is shown. In order to input the reanalysis data which will be used in further steps for creating the final dataset, the researcher has to select the option \textcolor{blue}{\textit{Manage reanalysis data}}.

@@

			Now, it will be described how to obtain wave classification models using the final dataset previously created with SPAMDA. The modelling will be performed using WEKA as \textcolor{red}{SC and} ML tool, which can be opened through SPAMDA, as shown in Fig. \ref{fig:openigFinalDatasetWeka}. Nevertheless, as mentioned above, the researcher can create the final dataset in CSV format in order to use other ML tools, such as KEEL, Python or R, among others.

@@

			\textcolor{red}{Quizas deberiamos poner algun parrafo mas sobre las conclusiones de la experimentacion, mirar el paper mio de ``Significant wave height and energy flux range forecast with machine learning classifiers'' por si sacar algo. Y quizas deberiamos hacer la experimentacion tambien con el flujo de energia en forma de regresion, ya que el special issue se titula ``Soft Computing Techniques in Energy System''. ¿Esta el dataset en el repositorio?}

@@

		A new open source tool named SPAMDA with an user-friendly GUI for creating datasets using meteorological data from NDBC and NNRP has been presented in this work. These datasets, that will be ready to be used as input for ML techniques in prediction tasks (classification or regression), are easily obtained by means of the selection of different input parameters, such as predictive and objective variables, output discretisation or prediction horizon. As a result, the researchers will benefit from significant support when carrying out environmental modelling related to energy, atmospheric or oceanic studies, among others. Moreover, given that SPAMDA simplifies all the intermediate steps involved in the creation of datasets and manages the extensive casuistry of the data integration (such as entering the meteorological information, managing with the incomplete data, pre-processing tasks, the customisable matching process to merge the data and the preparation of the datasets according to the ML technique to use), it avoids errors and reduces the time needed. In this way, the researchers will be able to have more in-depth analysis, which could result in more complete conclusions about the issue under study.

@@

\authorcontributions{Conceptualization, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández, Manuel Dorado-Moreno, Pedro Antonio Gutiérrez and César Hervás-Martínez; Formal analysis, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández, Manuel Dorado-Moreno, Pedro Antonio Gutiérrez and César Hervás-Martínez; Funding acquisition, Pedro Antonio Gutiérrez and César Hervás-Martínez; Investigation, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández, Manuel Dorado-Moreno, Pedro Antonio Gutiérrez and César Hervás-Martínez; Methodology, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández, Manuel Dorado-Moreno, Pedro Antonio Gutiérrez and César Hervás-Martínez; Project administration, Pedro Antonio Gutiérrez and César Hervás-Martínez; Resources, Pedro Antonio Gutiérrez and César Hervás-Martínez; Software, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández and Manuel Dorado-Moreno; Supervision, Pedro Antonio Gutiérrez and César Hervás-Martínez; Validation, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández, Manuel Dorado-Moreno, Pedro Antonio Gutiérrez and César Hervás-Martínez; Writing – original draft, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández and Manuel Dorado-Moreno. \textcolor{red}{ESTO DE LAS CONTRIBUCIONES HAY QUE VER SI LO CAMBIAMOS PARA QUE NO SEA TAN LARGO Y CON TANTA MORRALLA Y COMO PONERLO PARA QUE TE PUEDA VENIR BIEN POR SI LO NECESITASES PARA EL MASTER}}

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{Building Suitable Datasets for Soft Computing and Machine Learning Techniques from Meteorological Data Integration: A case  study for predicting significant wave height}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Meteorological data are extensively used to perform environmental learning. Soft Computing (SC) and Machine Learning (ML) techniques, a valuable support in many research areas, require datasets containing information related to the topic under study, which are not always available in an appropriate format, and its preparation and pre-processing implies a lot of time and effort by the researchers. This paper presents a novel software tool with an user-friendly GUI to create datasets by means of management and data integration of meteorological observations from two well-known sources of information: the \textit{National Data Buoy Center} and the \textit{National Centers for Environmental Prediction and for Atmospheric Research Reanalysis Project}. Such datasets can be created using buoys and reanalysis data by customisable procedures, in terms of predictive and objective variables and temporal resolution. These datasets can be used by SC and ML methodologies for prediction tasks (classification or regression) that support improvement of sustainability energy production, design of production systems as WECs or environmental modelling, among others.}

% Keywords
\keyword{Environmental Prediction; Renewable Energy Resource Evaluation; Meteorological Data; Reanalysis Data; Marine Energy; Soft Computing}  % List three to ten pertinent keywords specific to the article, yet reasonably common within the subject discipline.

@@

		A better understanding of the environment is of vital importance for science, contributing not only to more efficient exploitation of natural resources but also to the development of new strategies aimed at its protection. In that sense, meteorological observations provide an essential and valuable source of information which is widely used by researchers to address environmental learning, comprehension, prediction and conservation in numerous oceanic and atmospheric studies of a wide variety of areas (e.g. energy, climate change, agriculture, etc.). Some specific examples of the diversity of fields in which meteorological data can be used in are, among others: estimation of global solar radiation based on sunshine duration \cite{SHAHRUKHANIS2019179}, directional analysis of sea storms \cite{LAFACE201545}, estimation of hybrid energy systems taking into account economic and environmental objectives\cite{Kumar2020}, wind power ramp events prediction \cite{DORADOMORENO2017428}, sea surface temperature prediction \cite{He2020}, study of the responses exhibited by plankton to fluid motions \cite{FUCHS2016109}, trends in solar radiation \cite{SILVA20101852} or simulation of extreme near shore sea conditions \cite{GOULDBY201415}. All these studies require a prior data collection and its adaptation to a specific format that allows the interpretation of them.

		Once quality and well-formed data are obtained, these can be used to extract information and build prediction models that explain the behavior of a certain problem. Although Soft Computing (SC) and Machine Learning (ML) techniques have the ability to handle uncertainty in data and are extensively used for modelling purposes, the real challenge in modelling studies is due to the inadequacy of data, since the adequacy of the models depend mainly on the the quality of the information used, so that if a researcher does not have quality data there will be no quality models.

		Continuing with this line, special purpose software is usually developed to help researchers to advance in their studies related to energy and environmental modelling, becoming a great support for decision-making in the exploitation and protection of the environment. In \cite{Neeraj2020}, a software package in R called ``ForecastTB'' is developed to compare the accuracy of different forecasting methods as related to the characteristics of a time series dataset, presenting the software as a stepping stone in ML automation modelling.
		%In \cite{CAMARGONOGUEIRA2016361}, a software tool for designing solar water heating systems is developed, which simulates different situations and finds the best technical and economical solution. 
		In \cite{LO2015293}, an integrated simulation tool for the optimum design of bifacial solar panel with reflectors is presented. This tool can also be used to analyse the performance of the solar cells. A framework for data integration of offshore wind farms is implemented in \cite{NGUYEN2013150} in order to facilitate data exchange and improve operation and maintenance practices. 
		In \cite{Roberta2020}, a new software tool named ``Storage LCA Tool`` is presented for comparing PCM - phase change materials- storage systems  with traditional systems that do not entail energy storage, being beneficial in order to support decision-making on energy concepts for buildings.
		%In \cite{SHEHADEH2020100644}, a load management application to maximize the used energy of thermal systems and reduce the costs and emissions is developed. 
		A risk assessment tool to improve safety standards and emergency management in onshore wind farms, is presented in \cite{ASTIASOGARCIA201648}. Raabe et al. \cite{RAABE2010213} developed two software tools, \textit{Model of Equilibrium of Bay Beaches} (MEPBAY) and \textit{Coastal Modelling System} (SMC), to support different operational levels of headland-bay beach in coastal engineering projects, and Motahhir et al. \cite{MOTAHHIR20199} developed an open hardware/software test bench for solar tracker.

		Marine energy prediction is currently a hot topic where meteorological data is used in. Marine Renewable Energy (MRE) is one of the most important renewable and sustainable energy sources available in our environment \cite{en12050787}, and it includes ocean thermal energy, marine tidal current energy and wave energy, among others. Its benefits and great potential \cite{ZEYRINGER20181281} make it one of the most relevant natural resources, playing a crucial role not only in the reduction of the emission of greenhouse gases but also in all other aspects involved in the difficult challenge of the transition to a low carbon footprint society \cite{en12091657, BREDE201344}. Wave energy exhibits a more stable power supply than wind energy and even solar energy. In recent years \textit{Wave Energy Converters} (WECs) \cite{FALCAO2010899} have been developed and widely installed to transform this wave energy into electricity, which can be injected into the electric network or supplied to existing offshore oil and gas platforms \cite{OLIVEIRAPINTO2019556} or seawater desalination plants \cite{FERNANDEZPRIETO2019546}, among others. WECs are mechanical devices that convert kinetic energy into electrical energy by means of either the vertical oscillation of waves or the linear motion of them. Nevertheless, waves are difficult to be characterised due to their stochastic nature, because of the influence of a large number of environmental factors that exert on them \cite{ochi1998}. As a consequence of this complexity, many aspects of WEC design, deployment and operation \cite{CROWLEY2018159, Abdelkhalik2016, 6898109} need a proper prediction of waves, in order to maximise the wave energy extraction \cite{en80910370}. For this purpose WECs use wave \textit{flux of energy} ($F_e$) which can be calculated from the two most important wave parameters in this regard: \textit{significant wave height} ($H_s$) and \textit{wave energy period} ($T_e$). Additionally, wave predictions \cite{en11010011} are also helpful for designing offshore structures \cite{CHATZIIOANNOU2017126}, operational works in the sea \cite{DALGIC2015211}, providing technical information to the coastal and coral reef planners in developing countries \cite{CALLAGHAN2018123}, etc.

		Currently, and as a support to traditional study procedures, SC and ML techniques \cite{Sang-Yong2014,Bishop:2006:PRM:1162264} are being widely used in numerous research fields related to classification, regression and optimisation tasks, obtaining significant improvements in the performance of the results \cite{Fi-John2019,Mosavi2019,GUO201816}. SC and ML methodologies can be used not only by experienced computer scientists but also by other researchers. For example, the well-known \textit{Waikato Environment for Knowledge Analysis} (WEKA) \cite{WEKA} software tool provides researchers with a wide collection of ML  algorithms. In this way, ML techniques have been already applied to tackle wave characterisation, accurately estimating $H_{s}$ and $T_{e}$ parameters \cite{DURANROSAL2017268, KUMAR2017605}, given that robustness of ML methods can tackle the previously explained difficulties in wave energy prediction. The problem is that, in order to apply these methods, it is essential to obtain datasets with relevant information about the issue under study, used to infer knowledge. Usually, these datasets are not publicly available in a friendly format, and their generation is the first step needed.

		The information to create these datasets related with MRE can be obtained from meteorological observations, but such information may be available in a inappropriate format and even contain missing values or measurements. Consequently, it is usually required to perform pre-processing tasks for improving the quality of the data, such as the replacement of missing values, outlier detection or data normalisation, among others. Furthermore, if more than one source of information is used to achieve a better characterisation of the problem under study \cite{JOHANSSON2015143, FERNANDEZ201544, Adams2010}, then a data integration process, denominated as the matching process in this document, has to be carried out by the researchers to manually create the datasets with the needed information. Given that such process is of great relevance and has an extensive casuistry, the present work has been specially focused on it. Moreover, depending on the subject and the SC and ML technique to be applied, or even if the researcher considers other factors in order to improve the results or have more in-depth conclusions, the datasets would have to be updated afterwards. In summary, many important details and different intermediate steps have to be considered when creating suitable datasets, specially when data integration is required, resulting in an extremely tedious task.

		The main purpose of this paper is to present a new open source tool for the creation of datasets integrated by meteorological variables from two sources of information. Given that the tool provides an user-friendly graphical interface, no knowledge in programming languages is needed. It also prevents researchers from performing the mentioned tedious work and greatly simplify all the steps involved in it, avoiding possible errors in the intermediate steps, at least as a preliminary study in certain areas where some kind of environmental prediction is needed. The meteorological data used by the tool come from two well-known sources of information: the \textit{National Oceanic and Atmospheric Administration} (NOAA) \textit{National Data Buoy Center} (NDBC) \cite{NOAA} and the \textit{National Centers for Environmental Prediction} (NCEP)/\textit{National Center for Atmospheric Research} (NCAR) \textit{Reanalysis Project} (NNRP or R1) \cite{Kalnay1996, Kistler2001}. The open source software tool presented in this work is named SPAMDA (Software for Pre-processing and Analysis of Meteorological DAta to build datasets). As SPAMDA performs all this data processing, it reduces the time involving these tasks and allows the researchers focus on the study of the meteorological aspects of the observations. The datasets obtained are ready to be used as input for SC and ML techniques in prediction tasks (classification or regression), although the researchers can use them for other purposes. These datasets contain one or more meteorological variables as inputs and one variable as target (variable to be predicted). The format of the generated datasets will be \textit{Attribute-Relation File Format} (ARFF) \cite{WEKA_ARFF}, which is the one used by WEKA. Besides, the datasets can also be generated in \textit{Comma-Separated Values} (CSV) format, enabling the researchers to use others tools.

		In order to address the problem previously discussed, meteorological data integration from NDBC and NNRP and the casuistry that it entails, SPAMDA offers to the researchers novelties and functionalities that will be detailed in Section \ref{sec:SPAMDA}, although some of them are briefly summarised below:

		%Up to now, and to the best of our knowledge, this is the first software tool \textcolor{red}{with GUI (Graphical User Interface) }addressing the problem previously discussed, integrating meteorological data from NDBC and NNRP \textcolor{red}{and the casuistry that it entails}. \textcolor{blue}{(desde ''Up to now... hasta ...entails lo quitaría)} The novelties and functionalities that SPAMDA offers to the researchers will be detailed in Section \ref{sec:SPAMDA}, although some of them are briefly summarised below \textcolor{red}{Esto hay que revisarlo}:
		\begin{itemize}
			%\item The generation of datasets becomes a very easy and customisable task, by means of the selection of different input parameters, such as predictive and objective variables, classification and regression ML techniques, output discretisation or prediction horizon, among others.
			\item The generation of datasets becomes a very easy and customisable task by means of the selection of different input parameters, such as predictive and objective variables, classification and regression, output discretisation (useful for ordinal regression) or prediction horizon, among others.
			\item The created datasets can be easily used by SC and ML tools.
			\item It provides information about the quality and quantity of the data.
			\item It makes the researcher focus on environmental modelling, without having to worry about the development of scripts or mechanical tasks, avoiding laborious pre-processing procedures, that imply a lot of time and effort in early stages of the research.
			\item It avoids possible researcher errors in the intermediate steps of the process, such as geographical coordinates conversion, missing values handling (dates or measurements not recorded) or different temporal resolution of the data collected, among others.
			\item It manages the extensive casuistry of data integration which can lead to incomplete datasets, described in Appendix \ref{app:AppendixA}.
			\item Possibility of selecting one or more reanalysis nodes near the localisation under study.
			\item It also includes some basic pre-processing tasks, such as normalisation and missing data recovery.
			\item It facilitates data management and well-organised storage of the datasets.
			\item Its modular design allows the implementation of new modules for managing meteorological data from others sources, benefiting future renewable energy and environmental research.
			\item It includes a user-friendly GUI, facilitating and greatly simplifying data management, and it is integrated with the Explorer environment of WEKA.
			\item It is multi-platform, and it can be used on any computer with Java regardless of the operating system.
		\end{itemize}

@@

				Data pre-processing prepares the raw data (intermediate datasets) to be able to be treated correctly by ML algorithms. In this way, the quality of data can be improved prior to the learning phase, by applying pre-processing tasks (filters). The result will be referenced as pre-processed datasets.

@@

				Nevertheless, further pre-processing tasks can be performed after obtaining the final datasets by means of the Explorer environment of WEKA or other tools.

@@

				The automatic integration of the data provided by the two sources of information described in Section \ref{sec:DataSources}, to merge and format such data, is denominated as the matching process in this document. Such process is one of the most powerful and remarkable features of this software tool due to its great relevance and extensive casuistry. In this sense, SPAMDA has been developed to provide great flexibility to the researchers.

				The matching procedure is performed using an intermediate or pre-processed dataset, which includes the measurements collected by a buoy from NDBC, and the needed reanalysis data files from NNRP. Note that SPAMDA is able to manage the NetCDF binary format for handling the information stored in the reanalysis files.

@@

							\item \textit{One (using weighted mean of the $N$ nearest reanalysis nodes)}: Only one final dataset will be created, which will contain the attributes (the selected one as output and the selected ones as inputs) of the intermediate or pre-processed dataset used, along with a weighted mean of each variable of the reanalysis data used (one per selected reanalysis data file). This weighted mean is obtained by SPAMDA and uses Eq. \ref{eq:Haversine} to obtain the distance from each reanalysis node to the location of the buoy. Once the distances have been calculated they are inverted and normalised as follows:							
								\begin{linenomath*}
									\begin{equation}
										w_i=\frac{d(p_0,p_i)}{\sum_{j=1}^{N} d(p_0,p_j)}, ~~i=1, \ldots, N.
										\label{eq:weightedMean}
									\end{equation}
								\end{linenomath*}

@@

							\item \textit{'N' (one per each reanalysis node)}: As many final datasets as the number of nearest $N$ reanalysis nodes configured by the researcher will be created. Therefore, each final dataset will contain the value of each reanalysis variable used of the nearest corresponding reanalysis node, along with the selected attributes of the intermediate or pre-processed dataset used. In this way, researchers can perform comparison studies depending on the reanalysis node considered, to achieve better performance for the problem under study.

@@

	%\section{Case study}\label{sec:CaseStudy}
	\section{A case study applied to Gulf of Alaska}\label{sec:CaseStudy}		
		This section will describe how the application works in a practical approach. To do so, an example showing how to create a fully processed dataset (final dataset), starting from the raw data, will be described. The objective of this final dataset is to be used with ML algorithms for environmental modelling, concretely, to classify waves in the Gulf of Alaska depending on their height. In this case, the problem will be tackled as a multi-class classification problem, given that any continuous variable can be discretised in different classes. Such waves modelling can be applied with different purposes, energy flux range forecasting, missing buoy data reconstruction, or extreme significant wave heights detection, among others.

@@

			%After gathering the information described above\footnote{Further instructions for downloading this data can be found in the user manual of the application.}, the researcher can open SPAMDA. In Figure \ref{fig:main_view}, the main window is shown. In order to input the reanalysis data which will be used in further steps for creating the final dataset, the researcher has to select the option \textcolor{blue}{\textit{Manage reanalysis data}}.

			After gathering the information described above, the researcher can open SPAMDA. In Figure \ref{fig:main_view}, the main window is shown. In order to input the reanalysis data which will be used in further steps for creating the final dataset, the researcher has to select the option \textcolor{blue}{\textit{Manage reanalysis data}}.

@@

			Now, it will be described how to obtain wave classification models using the final dataset previously created with SPAMDA. The modelling will be performed using WEKA as SC and ML tool, which can be opened through SPAMDA, as shown in Fig. \ref{fig:openigFinalDatasetWeka}. Nevertheless, as mentioned above, the researcher can create the final dataset in CSV format in order to use other ML tools, such as KEEL, Python or R, among others.

@@

@@

		A new open source tool named SPAMDA with an user-friendly GUI for creating datasets using meteorological data from NDBC and NNRP has been presented in this work. These datasets, that will be ready to be used as input for SC and ML techniques in prediction tasks (classification or regression), are easily obtained by means of the selection of different input parameters, such as predictive and objective variables, output discretisation or prediction horizon. As a result, the researchers will benefit from significant support when carrying out environmental modelling related to energy, atmospheric or oceanic studies, among others. Moreover, given that SPAMDA simplifies all the intermediate steps involved in the creation of datasets and manages the extensive casuistry of the data integration (such as entering the meteorological information, managing with the incomplete data, pre-processing tasks, the customisable matching process to merge the data and the preparation of the datasets according to the SC or ML technique to use), it avoids errors and reduces the time needed. In this way, the researchers will be able to have more in-depth analysis, which could result in more complete conclusions about the issue under study.

		The case study illustrates how SPAMDA can be used by the researchers in a practical approach for environmental modelling, concretely, to classify waves in the Gulf of Alaska depending on their height.

@@

\authorcontributions{Conceptualization, Formal analysis and Investigation, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández, Manuel Dorado-Moreno, Pedro Antonio Gutiérrez and César Hervás-Martínez; Funding acquisition, Project administration, Resources and Supervision, Pedro Antonio Gutiérrez and César Hervás-Martínez; Methodology, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández, Manuel Dorado-Moreno, Pedro Antonio Gutiérrez and César Hervás-Martínez; Software, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández and Manuel Dorado-Moreno; Validation, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández, Manuel Dorado-Moreno, Pedro Antonio Gutiérrez and César Hervás-Martínez; Writing – original draft, Antonio Manuel Gómez-Orellana, Juan Carlos Fernández and Manuel Dorado-Moreno.}
